---
title: "Session-3"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Programming in R

## Conditionals

In order to update our function to decide between saving or not, we need to write code that automatically decides between multiple options.
The tool R gives us for doing this is called a conditional statement, and looks like this:

```{r, results='hold'}
num <- 37
if (num > 100) {
  print("greater")
} else {
  print("not greater")
}
print("done")
```

The second line of this code uses an `if` statement to tell R that we want to make a choice.
If the following test is true, the body of the `if` (i.e., the lines in the curly braces underneath it) are executed.
If the test is false, the body of the `else` is executed instead.
Only one or the other is ever executed:

flowchat

In the example above, the test `num > 100` returns the value `FALSE`, which is why the code inside the `if` block was skipped and the code inside the `else` statement was run instead.

```{r}
num > 100
```

And as you likely guessed, the opposite of `FALSE` is `TRUE`.

```{r}
num < 100
```

Conditional statements don't have to include an `else`.
If there isn't one, R simply does nothing if the test is false:

```{r}
num <- 53
if (num < 100) {
  print("num is less than 100")
}
```

We can also chain several tests together when there are more than two options.
This makes it simple to write a function that returns the sign of a number:

```{r}
sign <- function(num) {
  if (num > 0) {
    return(1)
  } else if (num == 0) {
    return(0)
  } else {
    return(-1)
  }
}
sign(-3)
sign(0)
sign(2/3)
```

Note that when combining `else` and `if` in an `else if` statement, the `if` portion still requires a direct input condition.  This is never the case for the `else` statement alone, which is only executed if all other conditions go unsatisfied.
Note that the test for equality uses two equal signs, `==`.

> **Other Comparisons**
>
> Other tests include greater than or equal to (`>=`), less than or equal to
> (`<=`), and not equal to (`!=`).
{: .callout}
We can also combine tests.
Two ampersands, `&&`, symbolize "and".
Two vertical bars, `||`, symbolize "or".
`&&` is only true if both parts are true:

```{r}
if (1 > 0 && -1 > 0) {
    print("both parts are true")
} else {
  print("at least one part is not true")
}
```

while `||` is true if either part is true:

```{r}
if (1 > 0 || -1 > 0) {
    print("at least one part is true")
} else {
  print("neither part is true")
}
```

In this case, "either" means "either or both", not "either one or the other but not both".

## Looping

- R has two basic types of loop
    + a **`for`** loop: run some code on every value in a vector
    + a **`while`** loop: run some code while some condition is true (*hardly ever used!*)
    
### `for` 

```{r eval=FALSE}
for(i in 1:10) {
  print(i)
  }
```

### `while`

```{r eval=FALSE}
i <- 1
while(i <= 10 ) {
  print(i)
  i <- i + 1
  }
```

## Functions

If we only had one data set to analyze, it would probably be faster to load the file into a spreadsheet and use that to plot some simple statistics.
But we have twelve files to check, and may have more in the future.
In this lesson, we'll learn how to write a function so that we can repeat several operations with a single command.

### Defining a Function

Let's start by defining a function `fahrenheit_to_kelvin` that converts temperatures from Fahrenheit to Kelvin:

```{r}
fahrenheit_to_kelvin <- function(temp_F) {
  temp_K <- ((temp_F - 32) * (5 / 9)) + 273.15
  return(temp_K)
}
```

We define `fahrenheit_to_kelvin` by assigning it to the output of `function`.
The list of argument names are contained within parentheses.
Next, the body of the function--the statements that are executed when it runs--is contained within curly braces (`{}`).
The statements in the body are indented by two spaces, which makes the code easier to read but does not affect how the code operates.

When we call the function, the values we pass to it are assigned to those variables so that we can use them inside the function.
Inside the function, we use a return statement to send a result back to whoever asked for it.

> ## Automatic Returns
>
> In R, it is not necessary to include the return statement.
> R automatically returns whichever variable is on the last line of the body
> of the function. While in the learning phase, we will explicitly define the
> return statement.
Let's try running our function.
Calling our own function is no different from calling any other function:

```{r}
# freezing point of water
fahrenheit_to_kelvin(32)
# boiling point of water
fahrenheit_to_kelvin(212)
```

We've successfully called the function that we defined, and we have access to the value that we returned.

### Composing Functions

Now that we've seen how to turn Fahrenheit into Kelvin, it's easy to turn Kelvin into Celsius:

```{r}
kelvin_to_celsius <- function(temp_K) {
  temp_C <- temp_K - 273.15
  return(temp_C)
}
# absolute zero in Celsius
kelvin_to_celsius(0)
```

What about converting Fahrenheit to Celsius?
We could write out the formula, but we don't need to.
Instead, we can compose the two functions we have already created:

```{r}
fahrenheit_to_celsius <- function(temp_F) {
  temp_K <- fahrenheit_to_kelvin(temp_F)
  temp_C <- kelvin_to_celsius(temp_K)
  return(temp_C)
}
# freezing point of water in Celsius
fahrenheit_to_celsius(32.0)
```

This is our first taste of how larger programs are built: we define basic
operations, then combine them in ever-larger chunks to get the effect we want.
Real-life functions will usually be larger than the ones shown here--typically half a dozen to a few dozen lines--but they shouldn't ever be much longer than that, or the next person who reads it won't be able to understand what's going on.

> ## Nesting Functions
>
> This example showed the output of `fahrenheit_to_kelvin` assigned to `temp_K`, which
> is then passed to `kelvin_to_celsius` to get the final result. It is also possible
> to perform this calculation in one line of code, by "nesting" one function
> inside another, like so:
>
> ```{r chained-example}
> # freezing point of water in Celsius
> kelvin_to_celsius(fahrenheit_to_kelvin(32.0))
> ```
> ## Create a Function
>
> In the last lesson, we learned to **c**ombine elements into a vector using the `c` function,
> e.g. `x <- c("A", "B", "C")` creates a vector `x` with three elements.
> Furthermore, we can extend that vector again using `c`, e.g. `y <- c(x, "D")` creates a vector `y` with four elements.
> Write a function called `highlight` that takes two vectors as arguments, called
> `content` and `wrapper`, and returns a new vector that has the wrapper vector
> at the beginning and end of the content:
>
> ```{r, echo=-1}
> highlight <- function(content, wrapper) {
>    answer <- c(wrapper, content, wrapper)
>    return(answer)
> }
> best_practice <- c("Write", "programs", "for", "people", "not", "computers")
> asterisk <- "***"  # R interprets a variable with a single value as a vector
>                    # with one element.
> highlight(best_practice, asterisk)
> ```
>
> > ## Solution
> > ~~~
> > highlight <- function(content, wrapper) {
> >   answer <- c(wrapper, content, wrapper)
> >   return(answer)
> > }
> > ~~~
> > {: .language-r}
> {: .solution}
>
> If the variable `v` refers to a vector, then `v[1]` is the vector's first element and `v[length(v)]` is its last (the function `length` returns the number of elements in a vector).
> Write a function called `edges` that returns a vector made up of just the first and last elements of its input:
>
> ```{r, echo=-1}
> edges <- function(v) {
>   first <- v[1]
>    last <- v[length(v)]
>    answer <- c(first, last)
>    return(answer)
> }
> dry_principle <- c("Don't", "repeat", "yourself", "or", "others")
> edges(dry_principle)
> ```
>
> > ## Solution
> > ~~~
> > edges <- function(v) {
> >   first <- v[1]
> >    last <- v[length(v)]
> >    answer <- c(first, last)
> >    return(answer)
> > }
> > ~~~

