---
title: "Exercises"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = FALSE)
```

> Write the follwing programs with-out using built-in functions.

1. Program to find factors of a number.
```{r}
print_factors <- function(x) {
print(paste("The factors of",x,"are:"))
for(i in 1:x) {
if((x %% i) == 0) {
print(i)
}
}
}
print_factors(21)
```

2. program to find the Factorial of a Number. This should give warning message for negetive numbers and zero.

```{r factorial}
# take input from the user
#num = as.integer(readline(prompt="Enter a number: "))
do_factorial <- function(num){
factorial = 1
# check is the number is negative, positive or zero
if(num < 0) {
print("Sorry, factorial does not exist for negative numbers")
} else if(num == 0) {
print("The factorial of 0 is 1")
} else {
for(i in 1:num) {
factorial = factorial * i
}
print(paste("The factorial of", num ,"is",factorial))
}
}
do_factorial(-5)
do_factorial(0)
do_factorial(7)
```

3. Program to sort numbers in a numeric vector.

```{r sorting}
sorting <- function(x){
  for(i in 1:(length(x)-1)){
    for(j in (i+1):length(x)){
      if(x[i] > x[j]){
        x[c(i,j)] = x[c(j,i)] # +1 for this
      }
    }
  }
  x
}
v <- c(3,1,4,7,2,9)
print("before sorting")
v
print("after sorting")
sorting(v)
```

4. Program to convert decimal number into binary number using recursive function.

```{r}
convert_to_binary <- function(n) {
if(n > 1) {
convert_to_binary(as.integer(n/2))
}
cat(n %% 2)
}
convert_to_binary(67)
```

5. Program to display the Fibonacci sequence up to n-th term using recursive functions.
```{r}
recurse_fibonacci <- function(n) {
if(n <= 1) {
return(n)
} else {
return(recurse_fibonacci(n-1) + recurse_fibonacci(n-2))
}
}
# take input from the user
nterms = as.integer(readline(prompt="How many terms? "))
# check if the number of terms is valid
if(nterms <= 0) {
print("Plese enter a positive integer")
} else {
print("Fibonacci sequence:")
for(i in 0:(nterms-1)) {
print(recurse_fibonacci(i))
}
}
recurse_fibonacci(4)
```

6. Program to check Armstrong Number.
```{r}
# take input from the user
num = as.integer(readline(prompt="Enter a number: "))
# initialize sum
sum = 0
# find the sum of the cube of each digit
temp = num
while(temp > 0) {
digit = temp %% 10
sum = sum + (digit ^ 3)
temp = floor(temp / 10)
}
# display the result
if(num == sum) {
print(paste(num, "is an Armstrong number"))
} else {
print(paste(num, "is not an Armstrong number"))
}
```

7. Program to Check for Leap Year.
```{r}
# Program to check if the input year is a leap year or not
year = as.integer(readline(prompt="Enter a year: "))
if((year %% 4) == 0) {
if((year %% 100) == 0) {
if((year %% 400) == 0) {
print(paste(year,"is a leap year"))
} else {
print(paste(year,"is not a leap year"))
}
} else {
print(paste(year,"is a leap year"))
}
} else {
print(paste(year,"is not a leap year"))
}
```

8. Program to show Multiplication Table of a number.
```{r}
# R Program to find the multiplicationtable (from 1 to 10)
# take input from the user
num = as.integer(readline(prompt = "Enter a number: "))
# use for loop to iterate 10 times
for(i in 1:10) {
print(paste(num,'x', i, '=', num*i))
}
```

9. Program to print length of each word in a sentence.
```{r}
x <- "R programing is awaesome"
y <- strsplit(x, " ")
for(i in y[[1]]){
  print(paste(i ," is length of ", length(i)))
}
```

10. Program to count vowels in a word.
```{r}
count_vowels <- function(word){
  y <- strsplit(word, "")
  num_of_vowels =0
  for(i in y[[1]]){
    if (i %in% c("a", "e", "i", "o", "u")){
      num_of_vowels=num_of_vowels+1
    }
  }
  print(paste("number of vowels in the word", word, "is ", num_of_vowels))
}
count_vowels("apple")
```