---
title: "session-5"
---

### Aim { .session-start }
* Learning how to graph/plot in R and basic customizations.

#
#

## Graphics
One of the main reasons data analysts turn to R is for its strong graphic capabilities.

## Creating a Graph
```{r}
attach(mtcars)
plot(wt, mpg)
abline(lm(mpg~wt))
title("Regression of MPG on Weight")
```

The `plot( )` function opens a graph window and plots weight vs. miles per gallon.
The next line of code adds a regression line to this graph. The final line adds a title.

## Saving Graphs
You can save the graph in a variety of formats from the menu
File -> Save As.

You can also save the graph via code using one of the following functions.

| Function                    | Output to        |
|-----------------------------|------------------|
| pdf("mygraph.pdf")          | pdf file         |
| win.metafile("mygraph.wmf") | windows metafile |
| png("mygraph.png")          | png file         |
| jpeg("mygraph.jpg")         | jpeg file        |
| bmp("mygraph.bmp")          | bmp file         |
| postscript("mygraph.ps")    | postscript file  |

## Built-in plots functions
Few ploting functions comes built-in when R installed. These can help do diffrent type of plots. Such as
  * Scater Plots
  * Dot Plots
  * Bar Plots
  * Pie Charts
  * Box Plots
  
## Scatter Plots
```{r}
plot(wt, mpg, main="Scatterplot Example",
   xlab="Car Weight ", ylab="Miles Per Gallon ", pch=19)
```
  
### Dot Plots
```{r}
dotchart(mtcars$mpg,labels=row.names(mtcars),cex=.7,
   main="Gas Milage for Car Models",
   xlab="Miles Per Gallon")
```

### Bar Plots

#### Simple Bar Plot
```{r}
counts <- table(mtcars$gear)
barplot(counts, main="Car Distribution",
   xlab="Number of Gears")
```

#### Stacked Bar Plot
```{r}
counts <- table(mtcars$vs, mtcars$gear)
barplot(counts, main="Car Distribution by Gears and VS",
  xlab="Number of Gears", col=c("darkblue","red"),
  legend = rownames(counts))
```

#### Grouped Bar Plot
```{r}
counts <- table(mtcars$vs, mtcars$gear)
barplot(counts, main="Car Distribution by Gears and VS",
  xlab="Number of Gears", col=c("darkblue","red"),
  legend = rownames(counts), beside=TRUE)
```

### Pie Charts

```{r}
slices <- c(10, 12,4, 16, 8)
lbls <- c("US", "UK", "Australia", "Germany", "France")
pie(slices, labels = lbls, main="Pie Chart of Countries")
```

### Box Plots

#### Simple Boxplot
```{r}
boxplot(mpg~cyl,data=mtcars, main="Car Milage Data",
   xlab="Number of Cylinders", ylab="Miles Per Gallon")
```

#### Notched Boxplot
```{r}
boxplot(len~supp*dose, data=ToothGrowth, notch=TRUE,
  col=(c("gold","darkgreen")),
  main="Tooth Growth", xlab="Suppliment and Dose")
```

## Customization
Graphical parameters describes how to change a graph's symbols, fonts, colors, and lines. Axes and text describe how to customize a graph's axes, add reference lines, text annotations and a legend. Combining plots describes how to organize multiple plots into a single graph.

```{r}
# Legend Example
attach(mtcars)
boxplot(mpg~cyl, main="Milage by Car Weight",
   yaxt="n", xlab="Milage", horizontal=TRUE,
   col=terrain.colors(3))
legend("topright", inset=.05, title="Number of Cylinders",
   c("4","6","8"), fill=terrain.colors(3), horiz=TRUE)
```

## Advance Ploting
Refer the follwing link - https://www.statmethods.net/advgraphs/index.html

### Key Points { .session-end }
* Legends of any plot can be made using `legend()` just after the plot is made.