---
title: "Session-4"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, cache = T)
```

### Aim { .session-start }
* Learn about packages in R
* Use Bioconductor

# Packages

Packages are collection of several functions in R.

To use a perticular function from a package, it need to be downloaded and loaded.

## Install a package from CRAN
```{r, eval=FALSE}
# download and install 'ggplot2' package from CRAN
install.packages("ggplot2")
```

## Introduction to Bioconductor
Bioconductor is a collection of more than 1,500 packages for the statistical analysis and comprehension of high-throughput genomic data. Originally developed for microarrays, Bioconductor packages are now used in a wide range of analyses, including bulk and single-cell RNA-seq, ChIP seq, copy number analysis, microarray methylation and classic expression analysis, flow cytometry, and many other domains.

This session introduces the essential of Bioconductor package discovery, installation, and use.

## Use Biocondutor
Discovering, installing, and learning how to use Bioconductor packages.

The web site at https://bioconductor.org contains descriptions of all Bioconductor packages, as well as essential reference material for all levels of user.

### Install from Bioconductor
```{r, eval=FALSE}
install.packages("BiocManager")
BiocManager::install(c("rtracklayer", "GenomicRanges"))
```

# Application: RNASeq DGE Analysis

RNA sequencing (RNA-seq) has proven as a revolutionary tool since the time it has been introduced. The throughput, accuracy, and resolution of data produced with RNA-seq has been instrumental in the study of Transcriptomics in the last decade (Wang, Gerstein, and Snyder 2009). There are many applications of Transcriptomics data, today we going to discuss Differential Gene Expression (DGE) which is wildly used to study the expression pattern of genes in contrasting samples.

Most of the steps of RNA-seq analysis have become quite mature over the years, especially how to reach a read count table from raw fastq reads obtained from an Illumina sequencing run. We will demonstrate how to process the count table, make a case-control differential expression analysis, and do some downstream functional enrichment analysis.


## About the data

For this exercise purpose we will be using the RNA-seq count table from a colorectal cancer study (SRA: [SRP029880](https://trace.ncbi.nlm.nih.gov/Traces/sra/?study=SRP029880), collected from recount2 database), only protein coding genes were kept for the faster execution of the analysis. 

Get to know the data

* [RAW count data](data/SRP029880.raw_counts.tsv)
* [Metadata](data/SRP029880.colData.tsv)

## Download the data

```{r download_data, eval=F}
dir.create("data")
download.file(url = "https://github.com/bionivid-tech/r-workshop-sep-2020/raw/master/data/SRP029880.raw_counts.tsv", 
              destfile = "data/SRP029880.raw_counts.tsv")
download.file(url = "https://github.com/bionivid-tech/r-workshop-sep-2020/raw/master/data/SRP029880.colData.tsv",
              destfile = "data/SRP029880.colData.tsv")
```


## Get data into R

Reading tabular in R.

```{r get_data_T}
counts <- read.table("data/SRP029880.raw_counts.tsv", sep = "\t", header = T, row.names = 1)
head(counts)
metadata <- read.table("data/SRP029880.colData.tsv", sep = "\t", header = T, row.names = 1)
metadata
```


## Normalization the data

In RNA-Seq, there are different type of normalization methods (such as - CPM, FPKM/RPKM, TPM) and some other method/package specific (Such as in DESeq2 - VST). Today we will calculate CPM and RPKM manually.

### CPM

Counts per million (CPM) mapped reads are counts scaled by the number of fragments you sequenced (N) times one million. This unit is related to the FPKM without length normalization and a factor of 10^3 [Ref](https://haroldpimentel.wordpress.com/2014/05/08/what-the-fpkm-a-review-rna-seq-expression-units/)


```{r cpm}
library(edgeR)
raw_counts <- subset(counts, select = c(-width))
cpm <- cpm.default(raw_counts)
head(cpm)
```

### RPKM

Reads per kilobase of exon per million reads mapped (RPKM), or the more generic FPKM (substitute reads with fragments) are essentially the same thing. Contrary to some misconceptions, FPKM is not 2 * RPKM if you have paired-end reads. FPKM == RPKM if you have single-end reads, and saying RPKM when you have paired-end reads is just weird, so don’t do it :). [Ref](https://haroldpimentel.wordpress.com/2014/05/08/what-the-fpkm-a-review-rna-seq-expression-units/)


```{r rpkm}
# create a vector of gene lengths 
geneLengths <- as.vector(subset(counts, select = c(width)))
# compute rpkm 
rpkm <- apply(X = subset(counts, select = c(-width)), 
                         MARGIN = 2, 
                         FUN = function(x) 10^9 * x / geneLengths / sum(as.numeric(x)))
options(scipen=999) # to off scientific notion of numbers
rpkm <- do.call(cbind, rpkm)
options(scipen=0, digits=7) # on again back
  
head(rpkm)
```

## Quality check of samples

For quality check of samples we can do several studies such as PCA and Correlation.

### PCA

Principal Compound analysis on samples.

```{r pca}
#transpose the matrix 
cpm_t <- t(cpm)
# transform the counts to log2 scale 
cpm_log <- log2(cpm_t + 1)
# calculate pca
pca <- prcomp(cpm_log)
summary(pca)
```

#### Plot PCA

Replicates should cluster together.

knitr::opts_chunk$set(echo = TRUE, warning=F, cache = F)

```{r pca_plot}
library(ggplot2)
library(ggfortify)
sample_pca <- autoplot(pca, data = metadata, colour = 'group')
sample_pca
```

### Correlation

Sample to sample relation.

```{r cor}
cor <- cor(cpm)
knitr::kable(cor)
```

#### Correlation Plot

```{r cor_plot}
library(corrplot)
sample_cor <- corrplot(cor, order = 'hclust', addCoef.col = 'white', addrect = 2)
```

#### Sample heatmap

Sample heatmap with annotation

```{r cor_heatmap}
library(pheatmap)
sample_heatmap <- pheatmap(cor, annotation_col = metadata)
```

## DEG Analsysis

Make the data ready.

```{r count_data}
#remove the 'width' column
countdata <- as.matrix(subset(counts, select = c(-width)))
head(countdata)
```

```{r meta_data}
metadata
```

Load the data into DESeq2 Object and design the experiment

```{r deseq_matrix_F, eval=F}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = countdata, 
                              colData = metadata, 
                              design = ~ group) # experiment design
dds
```

```{r deseq_matrix_T, echo=F}
suppressPackageStartupMessages(library(DESeq2))
dds <- DESeqDataSetFromMatrix(countData = countdata, 
                              colData = metadata, 
                              design = ~ group)
dds
```

Remove genes with almost no information in any of the given samples.

```{r remove_genes}
#For each gene, we count the total number of reads for that gene in all samples 
#and remove those that don't have at least 1 read. 
dds <- dds[ rowSums(counts(dds)) > 1, ]
```

### Run the analysis

```{r run_deseq2}
dds <- DESeq(dds) # the only time taking step. 
dds
```

### Explore results

```{r results}
#compute the contrast for the 'group' variable where 'CTRL' samples are used as the control group. 
res = results(dds, contrast = c("group", 'CASE', 'CTRL'))
head(res)
```

```{r res_sort}
#sort results by increasing p-value
res <- res[order(res$pvalue),]
res
```


### MA plot

```{r ma_plot}
DESeq2::plotMA(object = dds, ylim = c(-5, 5))
```

### P-value distribution

```{r p_val_dist}
ggplot(data = as.data.frame(res), aes(x = pvalue)) + geom_histogram(bins = 100)
```

### PCA 

For deseq normalized 

```{r pca_norm}
rld <- rlog(dds)
plotPCA(rld, ntop = 500, intgroup = 'group')
```

### Box plot

For deseq normalized 

```{r box_norm}
par(mfrow = c(1, 2))
boxplot(log(countdata), 
        col = as.numeric(as.factor(metadata$group)), 
        main = 'Raw Counts')
boxplot(log(counts(dds, normalized = TRUE)), 
        col = as.numeric(as.factor(metadata$group)), 
        main = 'Normalized Counts')
```

### DGE Volcano plot

```{r dge_volcano}
library(EnhancedVolcano)
EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = "CASE vs CTRL",
    pCutoff = 0.05,
    FCcutoff = 2,
    pointSize = 3.0,
    labSize = 3.0)
```

### Extract results

Extract the DGE to make heatmap

```{r extrcat_res}
library(dplyr)
library(tibble)
library(magrittr)
res_up <- res %>% as.data.frame() %>% 
  rownames_to_column('gene') %>%
  filter(log2FoldChange >= 2) %>% 
  mutate(reg_type = "Up regulation") %>% 
  column_to_rownames('gene')
head(res_up)
res_down <- res %>% as.data.frame() %>% 
  rownames_to_column('gene') %>%
  filter(log2FoldChange <= -2) %>% 
  mutate(reg_type = "Down regulation") %>% 
  column_to_rownames('gene')
head(res_down)
res_filt <- rbind(res_up, res_down)
# get up and down genes
res_filt_genes <- rownames(res_filt)
res_filt_genes[1:10]
```

### DGE Heatmap plot

```{r}
vsd <- vst(dds)
vsd_count <- assay(vsd)
dge_heatmap <- pheatmap(vsd_count[res_filt_genes,],
                       col = gplots::greenred(75),
                       scale = "row",
                       show_rownames = FALSE , 
                       cluster_rows=TRUE, cluster_cols=TRUE, 
                       annotation_col= metadata)
dge_heatmap
```

## Downstream analaysis

### Gene Ontology

```{r, echo=F}
suppressPackageStartupMessages(library(clusterProfiler))
suppressPackageStartupMessages(library(org.Hs.eg.db))
```

```{r go}
library(clusterProfiler)
library(org.Hs.eg.db)
ego <- enrichGO(gene          = res_filt_genes,
                OrgDb         = org.Hs.eg.db,
                keyType       = "SYMBOL",
                ont           = "CC",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
knitr::kable(head(ego@result))
```

### Enrichment Plots

#### Barplot 

```{r go_plot_1}
library(enrichplot)
go_bar <-barplot(ego, showCategory=20)
go_bar
```

#### Dot plot

```{r}
go_dot <- dotplot(ego, showCategory=20)
go_dot
```

#### Enrichment Map

```{r}
go_emap <-emapplot(ego)
go_emap
```


# Save data

## Save results

```{r, eval=F}
write.table(res_filt, "DGE_results_filtered.tsv", sep = "\t")
```

## Combine plots

```{r}
library(cowplot)
all_go_plot <- plot_grid(sample_pca, go_emap, labels = c('A', 'B'))
all_go_plot
```


## Save plots

```{r, eval=F}
tiff("DGE_heatmap.tiff")
all_go_plot
dev.off()
```


### Key Points { .session-end }
* To install packages from Biocondutor `BiocManager::install(pkg_name)`